\bf What is the Best Way to Display Two Time Series to Read Association?

### 2016-01-20 paper/paper.Rnw
first two paragraphs of section 2

%In order to work with data, it first must be understood. Statistical inference requires hypotheses to be established prior to data collection, but often data is collected first.
%This is especially so today, for vast databases that have been assembled in the big data era, that now need the data scientist to unravel the meaning of the numbers. 
%Without preset hypotheses to test, the power of statistical inference is impotent, and without hypotheses the data analyst can stumble blindly trying to build up models of structure in the data. 
%To understand data requires good visualisation. This idea was formed as early as the 18th century, when William Playfair institutionalized the then revolutionary idea of graphing government and economic data. 
%Far easier than reading tables of numbers, these ideas were powerful, and by providing the basic building blocks for plotting statistical data, his graphic designs became a conduit for the communication of otherwise complex information \citep{Sachs}. Since then, advances in computing power have allowed statistical graphics to flourish, spearheaded by \citet{Tukey} in 1965 into the new domain of exploratory data analysis (EDA) \citep{Friendly}. 
%EDA can be thought of as a well understood \citep{Cleveland, Vanderplas} set of tools and techniques required to visualise information, to physically see what the data contains. In particular, it incorporates a free roaming approach, where the analyst is able to explore structure to find whatever relationships and structure exists within. 
%Critically, the analyst does not have to have any pre-conceived ideas or hypotheses -- they are not specifically looking for any one particular thing. EDA emphasises letting the data inform us and can lead to the discovery of otherwise unexpected relationships, many of which may seem to become completely obvious after discovery. 
%With the knowledge provided by EDA, ideas are generated about what relationships between variables may potentially exist. This then enables the analyst to use these new hypotheses upon which to apply classical inference and rigorously conduct tests with new data.
%EDA is also related to the field of model diagnostics (MD), where a model can be continuously refined through the visualisation of its fit, its residuals, and the interactions with its variables. Both EDA and MD follow the same framework: Visualise the data, look for patterns that suggest an underlying relationship, and if one is found implement it into the model then continue the exploration of the data. 
%It has long been thought that EDA and statistical inference were worlds apart, but recent work in \citet{Buja} and \citet{Majumder:2013} bridges this chasm. Framing a data plot as a test statistic, which when compared to plots of null data, places EDA into the statistical inference recipe. 
%The null hypothesis underlying a particular plot, is generically that there is no pattern, and particular types of plots implicitly regulate what “no  pattern” means. For example, a scatter plot of two variables is used to explore for some sort of association, so the implicit hypothesis is that there is no association between the two variables.
%The alternative hypothesis is that there is some association, although it is not required to specify precisely the type of association. 
%The plot of the data is placed in a lineup of plots of null data, data generated assuming that the null hypothesis is true. If an impartial observer asked to select the plot that is most different from the rest, does select the data plot this suggests there exists a pattern that is not the result of chance, and the null hypothesis is rejected. 
%For a scatter plot, this departure from the null, might be a single outlier, or few outliers, a non-linear pattern, or clusters, which the human eye detects as more different in this data plot than in any of the nulls. This is the reason why visualisation remains important today, human eyes can detect patterns which would not be detected mathematically.  But eyes need calibration, which the lineup protocol provides.
%On its own, with a single plot, because of random sampling in collecting data, it is easy for the analyst to imagine a pattern when no real structure exists in the population. Visual skills of an observer in EDA is prone to Type I error, where the null hypothesis is rejected when it is actually true, caused by the inherently random formation of patterns when visualised which are attributed to structure rather than chance. \citet{Daniel} warns against this, by providing 40 pages of plots from the null distribution, he encourages data analysts to understand the patterns that can be created from data without inherent structure. (This is what \citet{Buja} call the Rorschach protocol.) By being aware of what can appear in this type of data, the analyst should be more wary of claiming any visual feature they find is a true relationship. But this is not enough, even seasoned data analysts can be misled.  \citet{Diaconis} introduced the notion of 'magical thinking' which argues that people commonly suffer from the over-interpretation of randomness, particularly if it matches a pre-conception. If it suits their particular bias, an analyst may make false discoveries of structure and false rejections of the null hypothesis. Whilst using both can complement each other, the treatment of Type I error has led EDA and inferential statistics to be considered as very disparate pursuits. The mathematical rigor that governs classical inference relies on a framework that acknowledges and controls for the rate of Type I error.  Section 2 describes the visual inference protocols which put EDA more firmly into the rigorous framework of statistical inference.

@@

>>>>

%In order to work with data, it first must be understood. Statistical inference requires hypotheses to be established prior to data collection, but often data is collected first.
%This is especially so today, for vast databases that have been assembled in the big data era, that now need the data scientist to unravel the meaning of the numbers. 
%Without preset hypotheses to test, the power of statistical inference is impotent, and without hypotheses the data analyst can stumble blindly trying to build up models of structure in the data. 
%To understand data requires good visualisation. This idea was formed as early as the 18th century, when William Playfair institutionalized the then revolutionary idea of graphing government and economic data. 
%Far easier than reading tables of numbers, these ideas were powerful, and by providing the basic building blocks for plotting statistical data, his graphic designs became a conduit for the communication of otherwise complex information \citep{Sachs}. Since then, advances in computing power have allowed statistical graphics to flourish, spearheaded by \citet{Tukey} in 1965 into the new domain of exploratory data analysis (EDA) \citep{Friendly}. 
%EDA can be thought of as a well understood \citep{Cleveland, Vanderplas} set of tools and techniques required to visualise information, to physically see what the data contains. In particular, it incorporates a free roaming approach, where the analyst is able to explore structure to find whatever relationships and structure exists within. 
%Critically, the analyst does not have to have any pre-conceived ideas or hypotheses -- they are not specifically looking for any one particular thing. EDA emphasises letting the data inform us and can lead to the discovery of otherwise unexpected relationships, many of which may seem to become completely obvious after discovery. 
%With the knowledge provided by EDA, ideas are generated about what relationships between variables may potentially exist. This then enables the analyst to use these new hypotheses upon which to apply classical inference and rigorously conduct tests with new data.
%EDA is also related to the field of model diagnostics (MD), where a model can be continuously refined through the visualisation of its fit, its residuals, and the interactions with its variables. Both EDA and MD follow the same framework: Visualise the data, look for patterns that suggest an underlying relationship, and if one is found implement it into the model then continue the exploration of the data. 
% Section 2 describes the visual inference protocols which put EDA more firmly into the rigorous framework of statistical inference.

@@

It has long been thought that EDA and statistical inference were worlds apart, but this is no longer true. \citet{Buja} and \citet{Majumder:2013} bridges the chasm. Framing a data plot as a test statistic, which when compared to plots of null data, places EDA into the statistical inference recipe. Underlying any data plot, there is an implicit null hypothesis that there is ``no pattern''. Particular types of plots may regulate what no  pattern means. For example, a scatter plot is used to explore for association between the two variables, so the implicit null hypothesis is that there is no association between the two variables. The alternative hypothesis is that there is some association, although it is not required to specify precisely the type of association. The departure from the null might be a single outlier, or few outliers, a non-linear trend, or clusters. To test the null hypothesis the data plot is placed in a lineup with plots of null data, data generated from a scenario of no pattern. An impartial viewer is asked to select the plot that is the most different from the others. If they select the data plot as the most different, this counts as a rejection of the null hypothesis, and follow-up questions may determine the type of departure: clusters, outlier, nonlinearity in the case of a scatterplot. With multiple independent observers a $p$-value can be calculated to measure the statistical significance of the departure from the null. This process is called the ``lineup protocol''.

The lineup protocol equips exploratory visual methods with a rigorous procedure that enables statistical inference. This is important because visualization remains essential for data analysis because human eyes can detect patterns which would not be detected mathematically. The lineup protocol provides the calibration of the human eyes to prevent us from erroneously jumping to conclusions about magic patterns. Earlier than the protocol proscription, we can find preliminary ideas along these lines, for example \citet{Daniel} recommends looking at plots of data from a null distribution for exploring two factor design experimental data.  His paper provides 40 pages of plots from the null! He encourages data analysts to understand the patterns that might be seen in samples data drawn from populations without inherent structure. (\citet{Buja} calls this the Rorschach protocol.)

%which the human eye detects as more different in this data plot than in any of the nulls.
%The plot of the data is placed in a lineup of plots of null data, data generated assuming that the null hypothesis is true. If an impartial observer asked to select the plot that is most different from the rest, does select the data plot this suggests there exists a pattern that is not the result of chance, and the null hypothesis is rejected. 
%For a scatter plot, this departure from the null, might be a single outlier, or few outliers, a non-linear pattern, or clusters, which the human eye detects as more different in this data plot than in any of the nulls. This is the reason why visualisation remains important today, human eyes can detect patterns which would not be detected mathematically.  But eyes need calibration, which the lineup protocol provides.
%On its own, with a single plot, because of random sampling in collecting data, it is easy for the analyst to imagine a pattern when no real structure exists in the population. Visual skills of an observer in EDA is prone to Type I error, where the null hypothesis is rejected when it is actually true, caused by the inherently random formation of patterns when visualised which are attributed to structure rather than chance. \citet{Daniel} warns against this, by providing 40 pages of plots from the null distribution, he encourages data analysts to understand the patterns that can be created from data without inherent structure. (This is what \citet{Buja} call the Rorschach protocol.) By being aware of what can appear in this type of data, the analyst should be more wary of claiming any visual feature they find is a true relationship. But this is not enough, even seasoned data analysts can be misled.  \citet{Diaconis} introduced the notion of 'magical thinking' which argues that people commonly suffer from the over-interpretation of randomness, particularly if it matches a pre-conception. If it suits their particular bias, an analyst may make false discoveries of structure and false rejections of the null hypothesis. Whilst using both can complement each other, the treatment of Type I error has led EDA and inferential statistics to be considered as very disparate pursuits. The mathematical rigor that governs classical inference relies on a framework that acknowledges and controls for the rate of Type I error.
